One-at-a-time Parameter Perturbation Ensemble of the Community Land Model, version 5.1

### 2025-04-15 ms/ms.tex
moving figs to ms directory for james latex

\section{Introduction}
Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within the terrestrial components of Earth System Models. Understanding how these processes respond to rising CO$_2$ and even influence CO$_2$ concentration trajectories is a critical facet of climate change research. Land processes substantially influence climate directly through, for example, evapotranspiration \cite{zarakas2024}, and indirectly through carbon-climate feedbacks \cite{cox2000}.
Uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}. 
Making centennial-scale projections of the cumulative terrestrial carbon sink has been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2006,friedlingstein2014,arora2020}, and limited efficacy of emergent constraints \cite{liu2024}. A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}. Still, our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns \cite{worden2021,baldocchi2024}. The capability to efficiently ingest these data and improve simulation performance is especially valuable for actionable science \cite{cheng2023}. However, given increasingly comprehensive land modeling systems, and the wide array of observational products, several technical hurdles exist that hinder effective model development and calibration.

@@

Understanding the range of outcomes arising from the many plausible parameter combinations is a critical step in robust uncertainty quantification. Parameter sensitivity tests can be used to gauge parametric uncertainty, and a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \add{new citation} \cite{dagon2020,baker2022,mcneall2024}. 
One important application of PPEs has been assessing uncertainty in climate model projections \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through various approaches like history matching, \add{which does not require assumptions regarding parameter distributions} \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024}, or Bayesian methods \add{new citations} \cite{ziehn2012,fer2018,cleary2021}, \add{or using genetic algorithms} \cite{bastrikov2018}.
Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large PPEs \cite{qian2018}.
In a more general sense, PPEs yield a robust knowledge basis for understanding and working with a given climate model. This can aid in introducing new users to the model, or to help steer and facilitate model development.

The process of parameter estimation for the complex land models typically embedded within Earth System Models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that comprise the terrestrial biosphere (e.g, hydrology, snow, biogeochemistry, physiology, land management, etc.). The number of model parameters required to represent all the relevant processes is large and model simulations are relatively expensive, presenting a major barrier to robust \remove{objective} calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
Thorough assessment of the parametric sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}. It typically requires extensive computational time,  software engineering, and domain-specific expert scientific knowledge. Our work here tries to alleviate each of these challenges in turn, at least within the context of one land model. While we are not yet at the point of automated calibration, we present herein several innovations that will serve to make an automated calibration system more technically feasible.

In this project we create a large PPE with the Community Land Model 5.1 with biogeochemistry configuration (CLM5.1-BGC, hereafter CLM)\cite{lawrence2019}. We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters. Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization. The goal of this project is to systematize the parameter perturbation process within the CLM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across a wide range of land model processes. In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
In this paper we present that dataset, describe how it was produced, and survey some potential applications.
The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b,zarakas2024}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and \remove{uncertainties} \add{sensitivities} within CLM.

@@

\subsection{Sparse grid}
\label{sect:sg}
Another control on model cost is spatial resolution. Most global CLM simulations utilize nominal 1$^\circ$ resolution, which equates to about 20,000 land grid cells. To reduce computational costs parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}. As opposed to selecting a coarse grid resolution, here we apply an alternative sparse grid approach by using a clustering algorithm to achieve an alternative low resolution configuration. The goal is to more strategically distribute grid cells, opting for more grid cells in areas that are especially distinct or variable, and fewer in areas that can be easily approximated based on output elsewhere (e.g. neighboring grid cells). 
Multivariate spatio-temporal clustering (MVSC) has been utilized to extract patterns of climatological significance from climate model output \cite{hoffman2005} and applied to design a representativeness-based sampling network \cite{hoffman2013}. Instead of lowering resolution by coarsening a rectilinear grid, we used MVSC to strategically remove effectively redundant grid cells. We used k-means clustering to identify groups of grid cells with similar dynamics based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase. We selected one representative grid cell from each cluster to stand in for the entire cluster. The representative grid cell is whichever is located nearest the cluster centroid in climate space. The set of representative grid cells comprise a `sparse grid', which are used in lieu of a `coarse' grid. To recompose mapped output and compute global means, the output from the representative grid cell is substituted for all members of the cluster cohort.

@@

Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants are hard-coded and not amenable to systematic perturbation \cite{mendoza2015,cuntz2016}. For the purposes of this activity, we identified a broad set of 211 CLM-BGC parameters, and enabled their modification via the model parameter file. Several domains were not perturbed, including crops, methane emissions, emissions of biogenic volatile organic compounds and dust, urban properties, and river flow parameters. The crop model was not utilized as our initial focus is on unmanaged land. Of the 211 parameters, 145 are scalar-valued, with a single parameter value acting globally. The remaining 66 parameters include a plant functional type (PFT) dimension, with the potential for different parameter values for each unique plant functional type. \add{For any given PFT parameter, all PFTs were perturbed in tandem (i.e. all 16 PFTs to their maximum values and then to their minimum values), but with potentially different ranges for each PFT. Explicit parameter ranges for each parameter (and each PFT) are provided in a supplemental spreadsheet.} \remove{Not all parameters are currently utilizing this functionality, such as \textit{medlynintercept} where the default value is 100 $\mu$mol H$_2$O m$^{-2}$ s$^{-1}$ for all sixteen of CLM's natural vegetation PFTs.}

@@

Given the large parameter space, to conduct an initial analysis of the response of the model to parametric uncertainty, we decided to vary each parameter independently, exploring the impact of low and high values. To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter. In some cases literature values were directly utilized, but in many cases, expert judgment was used or else a default perturbation of $\pm$20\%. The parameter range spreadsheet is provided within the Supplementary Information. In some cases parameters could not or should not be perturbed independently, because they feature inherent co-dependence. In the case of nitrogen fixation costs, we opted to perturb the parameters independently, but also as a group (`KCN'), to enforce a change in overall nitrogen limitation in lieu of switching from one uptake pathway to another. We likewise opted to perturb soil hydraulics directly in two distinct ways, first by directly perturbing the hydraulic properties (e.g. saturated hydraulic conductivity) and also \remove{then also} in a parallel set of simulations perturbing the soil texture (percent sand, silt, and clay) \add{and allowing the soil hydraulic properties to respond through the pedotransfer functions}.

@@

In one case, a significant CLM bug was identified after the completion of our primary experiment (\url{https://github.com/ESCOMP/CTSM/issues/2120}). Namely that the values for two nitrogen uptake parameters had been transposed in \citeA{brzostek2014}. Resolving the issue resulted in an approximately 5\% decrease in gross primary production of (Supp Figure S4) in the control simulations, and also significantly altered the size of the various parameter effects. Any PPE should be seen as intrinsically connected to its underlying codebase (bugs and all), and, reflecting this, we did not rerun the rest of the PPE. Therefore the majority of simulations contain the erroneous parameter values, and may slightly overestimate GPP. \add{The interacting effect of this bug fix with all of the other parameter perturbations was not explored.} The bug has since been resolved in later model versions \add{and will be tested in future PPEs}.

@@

 We opted for six different forcing scenarios to understand the intersection of parameter effects with different forcing agents associated with climate change (Table \ref{tab:exps}). The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, as it is the default forcing data for CLM5 \cite{lawrence2019}. Our six forcing scenarios were: a 2010-era control climate, high and low CO$_2$, future and pre-industrial climate anomaly forcing, and \add{enhanced} nitrogen deposition (see Table \ref{tab:exps}).  We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings. Future and pre-industrial climate anomaly forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to a mean climate change signal. We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest (either 1850s or 2090s) and present day for six atmospheric forcing variables (temperature, humidity, precipitation, wind, longwave and shortwave radiation; see Supp Figure S5 for climatologies). The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, which was chosen to align with the existing CESM2 large ensemble \cite{rodgers2021}. Since completing the experiment we uncovered a software bug affecting the AF1855 and AF2095 simulations. Namely that climate anomalies were not applied to certain coastal gridcells (Supp Fig S6). In all this affected less than 3\% of land area, and did not significantly alter parameter rankings (Supp Fig S7).

@@

\subsection{Biome definitions}
\label{sect:whit}
In addition to analyzing parameter effects on various output variables globally, we were also interested in understanding parameter effects on different biomes. As such, we categorized each of the 400 sparse grid cells according to their Whittaker biome \cite{whittaker1970}, which delineates biomes based on average temperature and precipitation. Biomes were delineated based on the atmospheric forcing temperature and precipitation data averaged across 2005 to 2014 (Supp Figure S8). Overall there are nine vegetated biomes and an additional ice sheet biome, where no vegetation is present (Supp Figures S9,10). \add{These biomes are somewhat arbitrary, and the biomes would vary slightly based on the climatology period, but they provide some coarse delineation that we feel has meaningful ecological significance and have found useful for analyzing this ensemble.}

\section{Results}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{../figs/main/bar.png}
\caption{211 parameters were perturbed across the various domains of the land model.}
\label{fig:params}
\end{figure}
Given the very large size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the relatively expensive computational cost of standard model configurations, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations. Combining the SASU spinup approach with the sparse grid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}). There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM5.1 simulations, respectively, as compared to just 400 grid cells in the sparsegrid. Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for this experiment within 140 years (Supp Figure S1).

\begin{figure}[h]
\centering
\includegraphics[width=30pc]{../figs/main/sims.png}
\caption{The approximate number of simulations afforded by 1 million core-hours on the Cheyenne supercomputer for a range of CLM configurations. Configurations are labeled according to spin-up procedure (SASU or the standard 1500-year spinup) and horizontal resolution (`sg' signifies sparsegrid). The inset map shows the locations of the 400 sparse grid cells. See Section \ref{methods} for spin-up and sparsegrid details.}
\label{fig:sims}
\end{figure}

Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Given more clusters, the sparsegrid will generally provide a better approximation of output from the full grid. As one example, accurate reconstruction of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters (Figure \ref{fig:sg}).
Performance continues to improve with added clusters, but with marginal returns (Supp Figure S2).
We had hoped to see a clear point of diminishing returns across a wide range of variables as we increase the number of clusters. Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, convergence slows and presents some small amplitude oscillations (Supp Figure S2). We chose 400 clusters, because it provides satisfactory performance across a range of important metrics, while affording a sufficient number of simulations to perform the full experiment within our computational budget.
\begin{figure}[h]
\centering
\includegraphics[width=25pc]{../figs/main/sparsegrid_gpp.pdf}
\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last forty years of a transient CLM5.1 simulation. We opted for 400 clusters to balance computational cost against representativeness.}
\label{fig:sg}
\end{figure}

This model configuration allowed for over \remove{2000} \add{2500} simulations across our six forcing scenarios.
Overall, we found substantial impacts of parametric \remove{uncertainty} \add{sensitivity} on model behavior, in some cases exceeding the magnitude of scenario effects (Figure \ref{fig:ranges}). In the control scenario, the global sum of gross primary production (GPP) ranged from 103 to 158 PgC per year across all ensemble members, and net ecosystem production (NEP) ranged from 1.7 to 3.8 PgC per year. NEP was especially variable in the high CO$_2$ scenario, varying from 1.6 to 5.7 PgC per year. One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death. A perturbation of $\pm$20\% may exceed the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils. Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, including unforeseen tipping points, brittle parameterizations, and/or bugs.

A small fraction of parameters tends to explain a large amount of the ensemble variance for any given variable or metric (Figure \ref{fig:variance}). GPP, which is a complicated process, requires 20 parameters to explain 90\% of the control ensemble variance. Conversely, canopy evaporation is governed by fewer parameters.  Just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S3). In general, most of the parameter perturbations had a relatively small impact on any given output variable. Though this experiment encompasses 211 parameters, no single variable responds to all 211 parameters. In fact, the effective parameter space dimensionality for any given output variable tends to be much smaller than 211. While the principle of land models is that these are all interlinked via complex feedback processes, many parameters nonetheless have impacts that are mostly limited to their own domain.

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{../figs/main/ranges.png}
\caption{\add{Global annual mean values for the default simulations (circles) and ensemble ranges (bars) for} (a) gross primary production (GPP), (b) net ecosystem production (NEP), \add{(c) leaf area index (LAI), and (d) evapotranspiration (ET)} across six forcing scenarios. \add{(e) Global LAI across the full set of parameter perturbation simulations under the future climate scenario. Most simulations fall within 5\% of the default LAI, with a small number of parameters generating larger effects. One perturbation, reducing sand heat capacity by 20\% caused a very low LAI.} \remove{Circles mark the CLM5.1 default simulation, and the bars span the ensemble ranges generated by individual parameter perturbations.}}
\label{fig:ranges}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{../figs/main/variance.png}
\caption{Cumulative fraction of variance explained by the most influential parameters on gross primary production and canopy evaporation. Approximately 96\% of canopy evaporation variance can be explained by the ten most influential parameters, whereas 30 parameters are required to explain 96\% of GPP variance.}
\label{fig:variance}
\end{figure}

>>>>

\section{Introduction}
Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within the terrestrial components of Earth System Models. Understanding how these processes respond to rising CO$_2$ and even influence CO$_2$ concentration trajectories is a critical facet of climate change research. Land processes substantially influence climate directly through, for example, evapotranspiration \cite{zarakas2024}, and indirectly through carbon-climate feedbacks \cite{cox2000}.
Uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}. 
Making centennial-scale projections of the cumulative terrestrial carbon sink has been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2006,friedlingstein2014,arora2020}, and limited efficacy of emergent constraints \cite{liu2024}. Emissions trajectories and land management scenarios contribute some uncertainty, but the majority of this uncertainty is inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}. Still, our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns \cite{worden2021,baldocchi2024}. The capability to efficiently ingest these data and improve simulation performance with improved parameters is especially valuable for actionable science \cite{cheng2023}. However, given increasingly comprehensive land modeling systems, and the wide array of observational products, several technical hurdles exist that hinder effective model development and calibration.

@@

Understanding the range of outcomes arising from the many plausible parameter combinations is a critical step in robust uncertainty quantification. Parameter sensitivity tests can be used to gauge parametric uncertainty, and a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \cite{dagon2020,baker2022,mcneall2024}. 
One important application of PPEs has been assessing uncertainty in climate model projections \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through various approaches like history matching, which does not require assumptions regarding parameter distributions \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024}, or Bayesian methods \cite{ziehn2012,fer2018,cleary2021}, or using genetic algorithms \cite{bastrikov2018}.
Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large PPEs \cite{qian2018}.
In a more general sense, PPEs yield a robust knowledge basis for understanding and working with a given climate model. This can aid in introducing new users to the model, or to help steer and facilitate model development.

The process of parameter estimation for the complex land models typically embedded within Earth System Models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that comprise the terrestrial biosphere (e.g, hydrology, snow, biogeochemistry, physiology, land management, etc.). The number of model parameters required to represent all the relevant processes is large and model simulations are relatively expensive, presenting a major barrier to robust calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
Thorough assessment of the parametric sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}. It typically requires extensive computational time,  software engineering, and domain-specific expert scientific knowledge. Our work here tries to alleviate each of these challenges in turn, at least within the context of one land model. While we are not yet at the point of automated calibration, we present herein several innovations that will serve to make an automated calibration system more technically feasible.

In this project we create a large PPE with the Community Land Model 5.1 with biogeochemistry configuration (CLM5.1-BGC, hereafter CLM)\cite{lawrence2019}. We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters. Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization. The goal of this project is to systematize the parameter perturbation process within the CLM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across a wide range of land model processes. In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
In this paper we present that dataset, describe how it was produced, and survey some potential applications.
The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b,zarakas2024}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and sensitivities within CLM.

@@

\subsection{Sparse grid}
\label{sect:sg}
Another control on model cost is spatial resolution. Most global CLM simulations utilize nominal 1$^\circ$ resolution, which equates to about 20,000 land grid cells. To reduce computational costs parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}. As opposed to selecting a coarse grid resolution, here we apply an alternative sparse grid approach by using a clustering algorithm to achieve an alternative low resolution configuration. The goal is to more strategically distribute grid cells, opting for more grid cells in areas that are especially distinct or variable, and fewer in areas that can be easily approximated based on output elsewhere (e.g. neighboring grid cells). 
Multivariate spatio-temporal clustering (MVSC) has been utilized to extract patterns of climatological significance from climate model output \cite{hoffman2005} and applied to design a representativeness-based sampling network \cite{hoffman2013}. Instead of lowering resolution by coarsening a rectilinear grid, we used MVSC to strategically remove effectively redundant grid cells. We used k-means clustering to identify groups of grid cells with similar dynamics based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase. We selected one representative grid cell from each cluster to stand in for the entire cluster. The representative grid cell is whichever is located nearest the cluster centroid in climate space. The set of representative grid cells comprise a `sparse grid', which are used in lieu of a `coarse' grid. To recompose mapped output and compute global means, the output from the representative grid cell is substituted for all members of the cluster cohort. This naive interpolation will introduce errors that grow in relative magnitude as the domain of interest shrinks.

@@

Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants are hard-coded and not amenable to systematic perturbation \cite{mendoza2015,cuntz2016}. For the purposes of this activity, we identified a broad set of 211 CLM-BGC parameters, and enabled their modification via the model parameter file. Several domains were not perturbed, including crops, methane emissions, emissions of biogenic volatile organic compounds and dust, urban properties, and river flow parameters. The crop model was not utilized as our initial focus is on unmanaged land. Of the 211 parameters, 145 are scalar-valued, with a single parameter value acting globally. The remaining 66 parameters include a plant functional type (PFT) dimension, with the potential for different parameter values for each unique plant functional type. For any given PFT parameter, all PFTs were perturbed in tandem (i.e. all 16 PFTs to their maximum values and then to their minimum values), but with potentially different ranges for each PFT. Explicit parameter ranges for each parameter (and each PFT) are provided in a supplemental spreadsheet.

@@

Given the large parameter space, to conduct an initial analysis of the response of the model to parametric uncertainty, we decided to vary each parameter independently, exploring the impact of low and high values. To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter. In some cases literature values were directly utilized, but in many cases, expert judgment was used or else a default perturbation of $\pm$20\%. The parameter range spreadsheet is provided within the Supplementary Information.

In some cases parameters could not or should not be perturbed independently, because they feature inherent co-dependence. In the case of nitrogen fixation costs, we opted to perturb the parameters independently, but also as a group (`KCN'), to enforce a change in overall nitrogen limitation in lieu of switching from one uptake pathway to another. For the grouped simulations, the given set of parameters were perturbed in tandem, either to their minimum or maximum values. We likewise opted to perturb soil hydraulics directly in two distinct ways, first by directly perturbing the hydraulic properties (e.g. saturated hydraulic conductivity) and also in a parallel set of simulations perturbing the soil texture (percent sand, silt, and clay) and allowing the soil hydraulic properties to respond through the pedotransfer functions.
Future experiments could utilize a multifactor perturbation sampling strategy to resolve parameter interdependencies and tradeoffs, such as along major axes of ecosystem function \cite{migliavacca2021}.

@@

In one case, a significant CLM bug was identified after the completion of our primary experiment (\url{https://github.com/ESCOMP/CTSM/issues/2120}). Namely that the values for two nitrogen uptake parameters had been transposed in \citeA{brzostek2014}. Resolving the issue resulted in an approximately 5\% decrease in gross primary production of (Supp Figure S4) in the control simulations, and also significantly altered the size of the various parameter effects. Any PPE should be seen as intrinsically connected to its underlying codebase (bugs and all), and, reflecting this, we did not rerun the rest of the PPE. Therefore the majority of simulations contain the erroneous parameter values, and may slightly overestimate GPP. The interacting effect of this bug fix with all of the other parameter perturbations was not explored. The bug has since been resolved in later model versions and will be tested in future PPEs.

@@

 We opted for six different forcing scenarios to understand the intersection of parameter effects with different forcing agents associated with climate change (Table \ref{tab:exps}). The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, as it is the default forcing data for CLM5 \cite{lawrence2019}. Our six forcing scenarios were: a 2010-era control climate, high and low CO$_2$, future and pre-industrial climate anomaly forcing, and enhanced nitrogen deposition (see Table \ref{tab:exps}).  We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings. Future and pre-industrial climate anomaly forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to a mean climate change signal. We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest (either 1850s or 2090s) and present day for six atmospheric forcing variables (temperature, humidity, precipitation, wind, longwave and shortwave radiation; see Supp Figure S5 for climatologies). The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, which was chosen to align with the existing CESM2 large ensemble \cite{rodgers2021}. Since completing the experiment we uncovered a software bug affecting the AF1855 and AF2095 simulations. Namely that climate anomalies were not applied to certain coastal gridcells (Supp Fig S6). In all this affected less than 3\% of land area, and did not significantly alter parameter rankings (Supp Fig S7).

@@

\subsection{Biome definitions}
\label{sect:whit}
In addition to analyzing parameter effects on various output variables globally, we were also interested in understanding parameter effects on different biomes. As such, we categorized each of the 400 sparse grid cells according to their Whittaker biome \cite{whittaker1970}, which delineates biomes based on average temperature and precipitation. Biomes were delineated based on the atmospheric forcing temperature and precipitation data averaged across 2005 to 2014 (Supp Figure S8). Overall there are nine vegetated biomes and an additional ice sheet biome, where no vegetation is present (Supp Figures S9,10). These biomes are somewhat arbitrary, and the biomes would vary slightly based on the climatology period, but they provide some coarse delineation that we feel has meaningful ecological significance and have found useful for analyzing this ensemble.

\section{Results}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{bar.png}
\caption{211 parameters were perturbed across the various domains of the land model.}
\label{fig:params}
\end{figure}

\subsection{Computational cost}

Given the very large size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the relatively expensive computational cost of standard model configurations, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations. Combining the SASU spinup approach with the sparse grid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}). There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM5.1 simulations, respectively, as compared to just 400 grid cells in the sparsegrid. Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for this experiment within 140 years (Supp Figure S1).

\begin{figure}[h]
\centering
\includegraphics[width=30pc]{sims.png}
\caption{The approximate number of simulations afforded by 1 million core-hours on the Cheyenne supercomputer for a range of CLM configurations. Configurations are labeled according to spin-up procedure (SASU or the standard 1500-year spinup) and horizontal resolution (`sg' signifies sparsegrid). The inset map shows the locations of the 400 sparse grid cells. See Section \ref{methods} for spin-up and sparsegrid details.}
\label{fig:sims}
\end{figure}

Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Given more clusters, the sparsegrid will generally provide a better approximation of output from the full grid. As one example, accurate reconstruction of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters (Figure \ref{fig:sg}).
Performance continues to improve with added clusters, but with marginal returns (Supp Figure S2).
We had hoped to see a clear point of diminishing returns across a wide range of variables as we increase the number of clusters. Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, improvement slows but does not saturate (Supp Figure S2). We chose 400 clusters, because it provides satisfactory performance across a range of important metrics, while affording a sufficient number of simulations to perform the full experiment within our computational budget.
The main goal of the sparsegrid is to obtain a coarse understanding of parameter effects at low computational cost, which could later be improved by full grid simulations or alternative sparsegrids with more clusters.

\begin{figure}[h]
\centering
\includegraphics[width=25pc]{sparsegrid_gpp.pdf}
\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last forty years of a transient CLM5.1 simulation. We opted for 400 clusters to balance computational cost against representativeness.}
\label{fig:sg}
\end{figure}

This model configuration allowed for over 2500 simulations across our six forcing scenarios.
Overall, we found substantial impacts of parametric sensitivity on model behavior, in some cases exceeding the magnitude of scenario effects (Figure \ref{fig:ranges}). In the control scenario, the global sum of gross primary production (GPP) ranged from 103 to 158 PgC per year across all ensemble members, and net ecosystem production (NEP) ranged from 1.7 to 3.8 PgC per year. NEP was especially variable in the high CO$_2$ scenario, varying from 1.6 to 5.7 PgC per year. One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death. A perturbation of $\pm$20\% may exceed the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils. Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, including unforeseen tipping points, brittle parameterizations, and/or bugs.

\subsection{Main results}

A small fraction of parameters tends to explain a large amount of the ensemble variance for any given variable or metric (Figure \ref{fig:variance}). GPP, which is a complicated process, requires 20 parameters to explain 90\% of the control ensemble variance. Conversely, canopy evaporation is governed by fewer parameters.  Just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S11). In general, most of the parameter perturbations had a relatively small impact on any given output variable. Though this experiment encompasses 211 parameters, no single variable responds to all 211 parameters. In fact, the effective parameter space dimensionality for any given output variable tends to be much smaller than 211. While the principle of land models is that these are all interlinked via complex feedback processes, many parameters nonetheless have impacts that are mostly limited to their own domain.

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{ranges.png}
\caption{Global annual mean values for the default simulations (circles) and ensemble ranges (bars) for (a) gross primary production (GPP), (b) net ecosystem production (NEP), (c) leaf area index (LAI), and (d) evapotranspiration (ET) across six forcing scenarios. (e) Global LAI across the full set of parameter perturbation simulations under the future climate scenario. Most simulations fall within 5\% of the default LAI, with a small number of parameters generating larger effects. One perturbation, reducing sand heat capacity by 20\% caused a very low LAI. }
\label{fig:ranges}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{variance.png}
\caption{Cumulative fraction of variance explained by the most influential parameters on gross primary production and canopy evaporation. Approximately 96\% of canopy evaporation variance can be explained by the ten most influential parameters, whereas 30 parameters are required to explain 96\% of GPP variance.}
\label{fig:variance}
\end{figure}

\subsection{Comparing scenarios and biomes}

We opted to rank parameters by absolute parameter effect, given that perturbations while unequal in relative size, were defined with the same uncertainty range in mind.
We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps} (Figure \ref{fig:ranges}). This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control the response to forcing changes (CO$_2$, nitrogen deposition and climate). For example, the parameters that control present-day NEP differ  from the parameters that control NEP in the high CO$_2$ scenario (Figure \ref{fig:nep}a,b). NEP increases by approximately 50\% with the increase of CO$_2$ from 367 to 867 ppm when using the CLM5.1 default parameters, but can actually decrease with certain parameter settings (Figure \ref{fig:nep}c). One key parameter is \textit{tpuse\_sf}, which is a scalar perturbation factor influencing the triose phosphate limitations on photosynthesis. Triose phosphate limitation is difficult to quantify and has not been observed to significantly hinder photosynthesis under present-day CO$_2$ concentrations, but may play a larger role in the future \cite{lombardozzi2018,kumarathunge2019}. This parameter is not particularly influential on NEP at 367 ppm (ranked 39, not shown), but is third most influential when we increased CO$_2$ to 867 ppm.
